---
title: "R Notebook"
output: html_notebook
---


# packages
```{r}
library(tidyverse)
library(readxl)
library(lubridate)
library(ordinal)
library(emmeans)
library(lme4)
library(scales)

```

```{r}
###palette controls for plotting colors, controls all plots


# Define color palette for groups outside the plot (put here so you can control later)
my_day_colors <- c(
  "1" = "#1f78b4",   # Nice blue
  "2" = "#6a3d9a",   # Violet
  "3" = "#fb9a99"    # Pink
)

# 

```



Mode Function
```{r}
Mode <- function(x, na.rm = FALSE) {
  if(na.rm){
    x = x[!is.na(x)]
  }

  ux <- unique(x)
  return(ux[which.max(tabulate(match(x, ux)))])
}
```



```{r}

#read in the data


Roost_Data <- read_excel("Roost Data.xlsx", 
    sheet = "ABC Scores aligned TEMP", col_types = c("text", 
        "numeric", "text", "numeric", "text")) %>%
      mutate(
     # Convert from character to numeric       
    Time= as.numeric(Time),
    #make time read as time 
    Time = format(as.POSIXct(Time * 86400, origin = "1970-01-01",  tz="America/Panama"), "%H:%M")
   
  ) %>%
      
      #rename number of videos 
      rename(n_videos = `# of videos`) %>%
      mutate(n_videos = as.numeric(n_videos), 
             Score = as.numeric(Score)) %>%
      
            filter(!is.na(Score)) %>%#filter out rows with NAs
  
  # Separate Trial into treatments, roost, and day 
  separate(Trial, into = c("Roost", "Treatment", "Day"), sep = "-", remove = FALSE)  %>%
      # Make Treatments names better
      mutate(Treatment = case_match( Treatment, 
                                 "S" ~ "Silence", 
                                 "H" ~ "Humans Talking", 
                                 "T" ~ "Traffic Noise"),
     #Use this line to order 
             Treatment = factor(Treatment,
                              levels = c(
                                "Silence",
                                "Humans Talking",
                                "Traffic Noise"
                                     ),
                        ordered = TRUE)
                         ) %>%
      mutate(Day = as.numeric(Day))
      


# Read in trial attributes

Roost_attributes <- read_excel("Roost Data.xlsx")
# View(Roost_attributes)

#left join roost attributes to get environmental data for each trial


str(Roost_Data)
```
# Summarize data

```{r}


Roost_summary <- Roost_Data %>%
      group_by(Roost, Trial, Treatment, Day) %>%
      summarize(n_videos= Mode(n_videos, na.rm = TRUE), # # of videos
                median = median(Score), #Median score
                max = max(Score), 
                Mode = Mode(Score),  #Modal score
                sum_score = sum(Score)) %>%  #change in sum(n_videos) to average when this fills in 
      mutate(disturbance_index = sum_score/n_videos)  #Summed all the ordinal behaviors/ # of videos 

Roost_summary
str(Roost_summary)

Roost_summary_SE <- Roost_summary %>%
  group_by(Treatment, Day) %>%
  summarize(
    mean_disturbance = mean(disturbance_index, na.rm = TRUE),
    se_disturbance = sd(disturbance_index, na.rm = TRUE) / sqrt(n()),# Standard Error
    mean_mode = mean(Mode, na.rm = TRUE), 
    se_mode = sd(Mode, na.rm = TRUE) / sqrt(n()),
    max = mean(max),
    .groups = "drop"
  )
```



# Grouped bar plot for the responses to the three treatments over three days
```{r}

# Define color palette for groups outside the plot (put here so you can control later)

my_day_colors <- c(
  "1" = "#1f78b4",   # Nice blue
  "2" = "#6a3d9a",   # Violet
  "3" = "#fb9a99"    # Pink
)

#Make the text larger "Bat Disturbance Index"  "HUmans Talking" 
# Description of the index

# Movement index
ggplot(Roost_summary_SE, aes(x = Treatment, y = mean_disturbance, fill = Day)) +
  geom_col(position = position_dodge(width = 0.9)) +     # Make bar chart
  geom_errorbar(aes(ymin = mean_disturbance - se_disturbance,   # Add in error bats 
                    ymax = mean_disturbance + se_disturbance),
                position = position_dodge(width = 0.9),
                width = 0.2) +
  geom_point(data = Roost_summary,  # Add in points from each trial over top 
             aes(x = Treatment, y = disturbance_index),
             position = position_jitterdodge(jitter.width = 0.1, dodge.width = .9),
             alpha = 0.6,
             size = 2) +
  theme_minimal(base_size = 18) +  #make color theme
  scale_fill_manual(values = my_day_colors) +
  ylab("Bat Disturbance Index") +  # Add x axis label
  xlab("Sound Treatment")   # Add y axis label

ggsave("Movement_index_plot.pdf")

ggsave("Movement_index_plot_wide.pdf",
       width = 8, 
       height = 5, 
       units = "in")


#Mode
ggplot(Roost_summary_SE, aes(x = Treatment, y = max, fill = Day)) +
  geom_col(position = position_dodge(width = 0.9)) +     # Make bar chart
  geom_errorbar(aes(ymin = mean_disturbance - se_mode,   # Add in error bats 
                    ymax = mean_disturbance + se_mode),
                position = position_dodge(width = 0.9),
                width = 0.2) +
  geom_point(data = Roost_summary,  # Add in points from each trial over top 
             aes(x = Treatment, y = max),
             position = position_jitterdodge(jitter.width = 0.1, dodge.width = .9),
             alpha = 0.6,
             size = 2) +
  theme_minimal(base_size = 18) +  #make color theme
  scale_fill_manual(values = my_day_colors) +
  ylab("Mode") +  # Add x axis label
  xlab("Sound Treatment")   # Add y axis label


 




```




#make plot showing frequencies of each score (this isa  total, might be good to normalize between roosts, control for the differences between roosts )
```{r}
#y= frequency x = scores 

#palette for humans treatment
my_treatment_colors <- c(
  "Silence" = "#1dd3b0",  # Turquoise
  "Humans Talking"  = "#76c893",  # Soft Green
  "Traffic Noise" = "#ffd166"   # Warm Yellow
)


Roost_Data %>%
      mutate(Score = case_match(
      Score,
      1 ~ "Ear Twitches",
      2 ~ "Moved location - walking",
      3 ~ "1 or 2 Flying",
      4 ~ ">1/2 Flying",
      5 ~ "Abandoning roost",
      .default = NA_character_
    ),
    Score = factor(
      Score,
      levels = c(
        "Ear Twitches",
        "Moved location - walking",
        "1 or 2 Flying",
        ">1/2 Flying",
        "Abandoning roost"
      ),
      ordered = TRUE
    )
  ) %>%
     ggplot(aes(x = Score, fill = Treatment)) +
  geom_bar() +  # Use geom_bar for categorical/factor x-axis
  facet_wrap(~Treatment, dir = "v") +
  theme_minimal(base_size = 18) +
  scale_fill_manual(values = my_treatment_colors) +
  xlab("Behavior Score") +
  ylab("Total Frequency") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) 

ggsave("behavior_frequencies_plot.pdf")



#plot no legend (you can delete legend )

Roost_Data %>%
      mutate(Score = case_match(
      Score,
      1 ~ "Ear Twitches",
      2 ~ "Walking",
      3 ~ "1 or 2 Flying",
      4 ~ ">1/2 Flying",
      5 ~ "Abandoning roost",
      .default = NA_character_
    ),
    Score = factor(
      Score,
      levels = c(
        "Ear Twitches",
        "Walking",
        "1 or 2 Flying",
        ">1/2 Flying",
        "Abandoning roost"
      ),
      ordered = TRUE
    )
  ) %>%
     ggplot(aes(x = Score, fill = Treatment)) +
  geom_bar() +  # Use geom_bar for categorical/factor x-axis
  facet_wrap(~Treatment, dir = "v") +
  theme_minimal(base_size = 22) +
  scale_fill_manual(values = my_treatment_colors) +
  scale_x_discrete(labels = label_wrap(10)) +
  xlab("Behavior Score") +
  ylab("Total Frequency") +
  theme(axis.text.x = element_text(angle = 0), 
        legend.position = "none") 

ggsave("behavior_frequencies_plot_new.pdf", 
       width = 7, 
       height = 7,
       units = "in")
                                
ggsave("behavior_frequencies_plot_wide.pdf",
       width = 7, 
       height = 5, 
       units = "in")                      
                              
                                
                        
                              

```

# Do bats have different behavioral responses to the different treatments?
# Are there significant differences between treatment and day? 


```{r}

#prepare data for stats
Roost_summary_stats <- Roost_summary %>%
      mutate(
    Treatment = factor(Treatment, ordered = FALSE), 
    Treatment = relevel(Treatment, ref = "Silence")) # ordinal response
 

#what does data look like? 
str(Roost_summary_stats)

head(Roost_summary_stats)

#what is the distribution of disturbance index? 
hist(Roost_summary$disturbance_index)


#Full model
model <- lmer( disturbance_index ~ Treatment * Day + (1 | Roost), data = Roost_summary_stats)

# model with no interaction between Treatment and Day
model_no_interaction <- lmer( disturbance_index ~ Treatment + Day + (1 | Roost), data = Roost_summary_stats)

#null model, only random effect of treatment
model_null <-lmer( disturbance_index ~ 1 + (1 | Roost), data = Roost_summary_stats)

#model with only Day
model_day <-lmer( disturbance_index ~ Day + (1 | Roost), data = Roost_summary_stats)

#model with only Treatment
model_treatment  <-lmer( disturbance_index ~ Treatment + (1 | Roost), data = Roost_summary_stats)


# Is the interaction model better than one with no interaction? 
anova(model, model_no_interaction)  # Likelihood Ratio Test
#no, better to use no interaction model 

# Is model better than a null model? 
anova(model_null, model_no_interaction)
#not much, but a little

#is model with treatment alone better?
anova(model_null, model_treatment)
#no

anova(model_null, model_day)
#yes: day is better  than no interaction 

anova(model_no_interaction, model_day)
#no

# what does this model say: 
summary(model_no_interaction)

model_ni_emmmeans <- emmeans(model_no_interaction , ~ Treatment)
model_ni_emmmeans
plot(model_ni_emmmeans)
ggsave("emmeans_estimates.jpg")

model_ni_emmeans_day <- emmeans(model_no_interaction , ~ Day )  # Treatment effects within each Day
model_ni_emmeans_day
plot(model_ni_emmeans_day)

# or


 plot(model_ni_emmeans_2)
 
 
 # Explanation of model: 
#Response variable: disturbance_index

#Fixed effects: Treatment and Day

#Random effect: random intercept for Roost (i.e., you're allowing for variation in baseline disturbance_index across different roosts)
# Roost random intercept: There’s noticeable variability across roosts (SD = 1.743), suggesting that roost identity accounts for some variation in disturbance index.
# 
# Residual SD: The remaining within-roost variation is larger (SD = 2.484), meaning most of the variation is within roosts over time or across treatments.

```








```{r}

#################

# 
# #what is the distribution of the scores? 
# hist(Roost_Data$Score)
# # data have a  of weird distribution
# 
# 
# # Possible Model : Score ~ treatment*day (1|roost)
# 
# #Format data for analysis
# Roost_Data_Stats <- Roost_Data %>%
#   mutate(
#     Score = factor(Score, ordered = TRUE),  # ordinal response
#     Day =  as.numeric(Day) ,#factor(Day, ordered = TRUE),                      # ordinal predictor
#     Roost = factor(Roost, ordered = FALSE),                   # grouping factor
#     Treatment = factor(Treatment)
#   ) 
# 
# 
# str(Roost_Data_Stats)
# 
# head(Roost_Data_Stats)
# hist(Roost_Data_Stats$Score)
# 
# #Full model: model if score is responsive to treatment by day
#   model <- glmer(Score ~ Treatment * Day + (1 | Roost), data = Roost_Data_Stats, family = poisson())
# summary(model)
# 
# # model with no interaction 
# model_no_interaction <- glmer(Score ~ Treatment + Day + (1 | Roost), data = Roost_Data_Stats)
# 
# #null model, only random effect of treatment
# model_null <- glmer(Score ~ (1|Roost), data = Roost_Data_Stats)
# 
# 
# # Is model better than a null model? 
# anova(model, model_null)
# anova(model_null, model_no_interaction)
# anova(model, model_no_interaction)  # Likelihood Ratio Test
# 
# 
# emmeans(model, ~ Treatment)
# 
# 
# emmeans(model, ~ Treatment | Day)  # Treatment effects within each Day
# # or
# emmeans(model, ~ Day | Treatment)  # Day effects within each Treatment
# 
# plot(emmeans(model, ~ Treatment | Day, mode = "prob"))


###########



```



